Generalizing the Model View Controller Paradigm

### 2006-07-30 inst/paper/MVC.tex
did some commits
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18999 bc3139a8-67e5-0310-9ffc-ced21a209358

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\cite{intGrUnwin} \cite{iPlots} \cite{MONDRIAN} \cite{GGobi}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.

We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked models.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell model B to update.  Then all views of model B must be
      updated.  Next any data that are linked to model B must be notified
      that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, such as the data shown in
panel (a) of Figure~\ref{Fig:OneDBTab}.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (panel (b) of Figure~\ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

@@

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a matrix. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  Even though there are two \Robject{MVC} objects
here, one \Robject{MVC} for each model, the views are linked.

@@

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that are
needed by all models.  Then, we created a number of subclasses, one for
each type of model.  Currently, we have implemented three types of models, 
one based on data frames, one based on graphs, and one based on the 
expression set class of data used by the Bioconductor project \citep{BIOC}.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own package.

@@

As for specific view classes, the \Rclass{spreadView} class represents
a spreadsheet view of the data.  This view only makes sense for models
that are a two dimensional data structure, such as a matrix or a data
frame. Other views we have implemented include: scatter plots
(\Rclass{sPlotView}), graphs (\Rclass{graphView}), and heatmaps
(\Rclass{heatmapView}).

@@

The \Rclass{MVC} class binds the model, view, and controller objects
into one object.  Thus, whenever a new data set is loaded, an
appropriate \Robject{MVC} object is created.  There is a one-to-one
relationship between the \Robject{MVC} instances and model objects.
Because of this relationship, the \Robject{MVC} object and the model
object are identified by the same name.

Our implementation consists
of a virtual class, \Rclass{MVC}, with three slots, which are
\Rslot{model}, \Rslot{viewList}, and \Rslot{controller}. The three
slots correspond to the three pieces of information that are needed to
bind the model, view, and controller into a single unit. This class is
sufficient for all single MVC activities. To support linked models
we extend this class with the \Rclass{linkedModelMVC} class, which has
two additional slots: \Rslot{parentMVC}, which is the name of the
parent model and \Rslot{childMVCList}, which is a list of the child models.
These are the models that will communicate with each other: parents send
information to their children and children send information to their parent.
More information on this communication is given in Section \ref{Sec:MultMVC}.

@@

%%FIXME: at this point we could perhaps even include a real example
%%with pictures of what is described below?
%% EAW: Yes, we could.  My only concern is confusing the idea of views
%% with models.  I think it's important to show the inheritance
%% structure of the models and not just views of the models.  
Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects are related to each other through a
parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
picture, `modelA' is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
`modelB' and `modelC'. Then from `modelB', three
\Robject{MVC} objects were created, `modelD', `modelE', and
`modelF'.  As an example, modelA could be gene 
expression data, modelB could be gene expression data that is a subset of
modelA (based on filtering), modelC could be derived by performing
multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
(GO) graph representing the molecular functions of the genes in modelB,
modelE could be the KEGG pathways that the genes in modelB are involved
in, and modelF could be the chromosome locations of the genes in modelB.

@@

Deciding when new data should be added to the current model versus generating
a new submodel in a \Robject{MVC} instance is done by comparing the
relationship between the new data and the data in the current model.  If the
new data has a one-to-one relationship with the elements in the current model,
then the new data can be added to the current model.  For example, if there is
a test statistic for each row in a data frame, then this could be added to the
model.  However, if the new data has a one-to-many or a many-to-many
relationship with the elements in the current model, then a submodel is
generated in a new \Robject{MVC} instance.  An example here is the
relationship between genes and GO terms.  If the original model contains gene
expression levels and a user wants to explore the GO terms for these genes,
then the GO graph representing the GO terms will be placed in a new
\Robject{MVC} because the genes and GO terms have a many-to-many relationship.

@@

%%FIXME: if it is not too much work, lets give the models names, not
%%numbers.
%% EAW: It's been changed.
Consider Figure~\ref{Fig:Hier}, which was shown previously.  If
modelA's model changes, then this information first gets passed to
modelB and modelC and these two \Robject{MVCs} may make changes to
their models based on this information.  Then modelB passes this
information to modelD, modelE, and modelF based on the changes
that modelB made to its model.  Thus, information passing between
modelA and modelD, for instance, only occurs through modelB.  So
messages can only be sent from parent \Robject{MVC} to child
\Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
This communication between parent and child \Robject{MVCs} was shown in Figures
\ref{Fig:heatmap}, \ref{Fig:GOgraph}, and \ref{Fig:heatmapsub} in Section
\ref{sec:micro}.  Here views of the parent and child models are
synchronized, but the selected items in a grandparent model may not
match the selected items in a grandchild model (as shown by Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} where the selected genes are
different in these two heatmaps).  Determining how \Robject{MVC} models are
linked is a difficult task when the relationship between elements in the two
models is not one-to-one.

@@

The goals for this paper were to create linked views of linked data sets, to
create interactive views where the response to an event can be changed, and to
create an extensible design that allows future users to make additions.
The first goal was accomplished by implementing \Robject{MVC} objects as
components that communicate with each other through messages.  Encapsulating
each data set in a \Robject{MVC} object was particularly useful because it
allowed the reuse of code for communicating between the components of a
\Robject{MVC} object, which made generating new model and view classes easier.
However, communication between the \Robject{MVC} objects was, at times, more
complicated.

@@

The functionality that is defined in the \Rpackage{iSNetwork} package is
discussed in detail in the Vignette for this R package.  Users can load data,
create views of the data, interact with these views and change the response to
the interaction (these interactions include mouse button click events and
mouse movement events), and users can create child \Robject{MVC} objects
through the GUI or through command line functions.  With the functionality in
the \Rpackage{iSNetwork} package, the user can create views of linked models
and can explore visually the relationships between variables in different data
sets.  This functionality was implemented by extending the MVC paradigm in a
very natural way to encompass a much richer class of models and user
interactions.

>>>>

%%Cite DataDesk as a system that does this, 
%% also this book:  Andreas Buja, Catherine Hurley, and John Alan McDonald. Elements of a viewing pipeline for data analysis. In W. S. Cleveland and M. E. McGill, editors, Dynamic Graphics for Statistics. Wadsworth and Brooks/Cole, Belmont, Ca., 1988.

%% also this paper: Interactive data visualization using focusing and linking
%%Full text 	pdf formatPdf (541 KB)
%%Source 	IEEE Visualization archive
%%Proceedings of the 2nd conference on Visualization '91 table of contents
%%San Diego, California
%%SESSION: Multivariate visualization table of contents
%%Pages: 156 - 163  
%%Year of Publication: 1991
%%ISBN:0-8186-2245-8 (PAPER) 
There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data.  Many authors
\citep{intGrUnwin, iPlots, MONDRIAN, GGobi}
have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data.  The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems.  The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.
%%FIXME: cite the JSS submission
We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked models.  When a user interacts with
      the scatter plot, the first step, indicated by a circled 1,
      is to tell model B to update.  Then all views of model B must be
      updated.  Next any data that are linked to model B must be notified
      that model B changed.  This information from model B must be
      altered so that it can be understood by linked models, which is shown
      in the fourth step.  Finally, model A is notified that it should
      update in response to the change in model B and then all views of
      model A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, such as the data shown in
Figure~\ref{Fig:DBTab}.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

@@

We felt that neither approach had much appeal.  A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlighted.  The second approach requires writing a
large amount of code, but is unlikely to be extensible.
We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a rectangular array. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

@@

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  But there are two distinct MVCs, not one. 
The city data have one MVC and the state data a second.  Even though 
there are two \Robject{MVC} objects here the views are linked.

@@

In our MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that are
needed by all models.  Then, we created a number of subclasses, one for
each type of model.  Currently, we have implemented three types of models, 
one based on data frames, one based on graphs, and one based on the 
expression set class of data used by the Bioconductor project \citep{BIOC}.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own R software package.

@@

As for specific view classes, the \Rclass{spreadView} class represents
a spreadsheet view of the data.  This view requires input data that are
inherently two-dimensional and for graph data, we make a reduction to the 
list of nodes and some attributes.  Other views we have implemented 
include: scatter plots
(\Rclass{sPlotView}), graphs (\Rclass{graphView}), and heatmaps
(\Rclass{heatmapView}).

@@

The \Rclass{MVC} class binds the model, view, and controller objects
into one object.  Thus, whenever a new data set is loaded, an
appropriate \Robject{MVC} object is created.  There is a one-to-one
relationship between the \Robject{MVC} instances and model objects.
%%FIXME: this seems like an implementation detail
%%Because of this relationship, the \Robject{MVC} object and the model
%%object are identified by the same name.
Our implementation consists
of a virtual class, \Rclass{MVC}, with three slots, which are
\Rslot{model}, \Rslot{viewList}, and \Rslot{controller}. The three
slots correspond to the three pieces of information that are needed to
bind the model, view, and controller into a single unit. This class is
sufficient for all single MVC activities. To support linked models
we extend this class with the \Rclass{linkedModelMVC} class, which has
two additional slots: \Rslot{parentMVC}, which is the name of the
parent model and \Rslot{childMVCList}, which is a list of the child models.
These are the models that will communicate with each other: parents send
information to their children and children send information to their parent.
More information on this communication is given in Section \ref{Sec:MultMVC}.

@@

%%FIXME: at this point we could perhaps even include a real example
%%with pictures of what is described below?
%% EAW: Yes, we could.  My only concern is confusing the idea of views
%% with models.  I think it's important to show the inheritance
%% structure of the models and not just views of the models.  
Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects can be related to each other 
through a
parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
picture, modelA is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
modelB and modelC. Then from `modelB', three
\Robject{MVC} objects were created, modelD, modelE, and
modelF.  As an example, modelA could be gene 
expression data, modelB could be gene expression data that is a subset of
modelA (based on filtering), modelC could be derived by performing
multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
(GO) graph representing the molecular functions of the genes in modelB,
modelE could be the KEGG pathways that the genes in modelB are involved
in, and modelF could be the chromosome locations of the genes in modelB.

@@

Deciding when new data should be added to the current model versus generating
a new submodel in a \Robject{MVC} instance is done by comparing the
relationship between the new data and the data in the current model.  If the
new data has a one-to-one relationship with the elements in the current model,
then we believe that in most cases the new data should be added to the 
current model.  For example, if there is
a test statistic for each row in a data frame, then this could be added to the
model.  However, if the new data has a one-to-many or a many-to-many
relationship with the elements in the current model, then serious consideration should be given to the creation of a submodel.  If there is already an appropriate MVC class (e.g. if the submodel is a graph) then it can be
generated by creating a new instance of the appropriate MVC class.  If there
is no appropriate implementation, then the user will need to create one.  
As an example, consider the
relationship between genes and GO terms.  If the original model contains gene
expression levels and a user wants to explore the GO terms for these genes,
then we encourage the creation of the GO graph representing the GO terms,  and
creating a new graph \Robject{MVC} object.
%%FIXME: maybe explicitly name the class in the above line

@@

%%FIXME: if it is not too much work, lets give the models names, not
%%numbers.
%% EAW: It's been changed.
Consider Figure~\ref{Fig:Hier}, which was shown previously.  If
modelA's model changes, then this information first gets passed to
modelB and modelC and these two \Robject{MVCs} may make changes to
their models based on this information.  Then modelB passes this
information to modelD, modelE, and modelF based on the changes
that modelB made to its model.  Thus, information passing between
modelA and modelD, for instance, only occurs through modelB.  So
messages can only be sent from parent \Robject{MVC} to child
\Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
This communication between parent and child \Robject{MVCs} was shown in Figures
\ref{Fig:heatmap}, \ref{Fig:GOgraph}, and \ref{Fig:heatmapsub} in Section
\ref{sec:micro}.  Here views of the parent and child models are
synchronized, but the selected items in a grandparent model may not
match the selected items in a grandchild model (as shown by Figures
\ref{Fig:heatmap} and \ref{Fig:heatmapsub} where the selected genes are
different in these two heatmaps).  Deciding how to link \Robject{MVC} models 
is a difficult task especially when the relationship between elements in the 
two models is not one-to-one.

@@

The goals for this research were to create software implementations and conceptual paradigms to support the use of linked views of linked data sets, to
create interactive views where the response to an event can be changed, and to
create an extensible design that allows future users to make additions.
The first goal was accomplished by implementing \Robject{MVC} objects as
components that communicate with each other through messages.  Encapsulating
each data set in a \Robject{MVC} object was particularly useful because it
allowed the reuse of code for communicating between the components of a
\Robject{MVC} object, which made generating new model and view classes easier.
However, communication between the \Robject{MVC} objects was, at times, more
complicated.

@@

The functionality that is provided in the \Rpackage{iSNetwork} package is
discussed in detail in the Vignette for this R package.  Users can load data,
create views of the data, interact with these views and change the response to
the interaction (these interactions include mouse button click events and
mouse movement events), and users can create child \Robject{MVC} objects
through the GUI or through command line functions.  With the functionality in
the \Rpackage{iSNetwork} package, the user can create views of linked models
and can explore visually the relationships between variables in different data
sets.  This functionality was implemented by extending the MVC paradigm in a
very natural way to encompass a much richer class of models and user
interactions.
